---
title: "Time spent travelling to the disposal site during pit emptying jobs in Mzuzu, Malawi"
author:
  - name: Padraic Casserly
    email: pcasserly@ethz.ch
    affiliations:
      - name: Global Health Engineering, ETH Zürich
date: 2026-09-29
bibliography: references.bib
abstract: |
  We use task-level timing data recorded with the Gulper Timer app during
  ridealong observations of manual pit latrine emptying in Mzuzu, Malawi, to
  quantify how much time a pit emptying team spends travelling to the sludge
  disposal site, for the jobs in which such a trip was made. Travel time is
  reported per job in two ways: elapsed time (clock hours during which the
  team was on a trip, with simultaneous travel by several workers counted
  once) and labour time (person-hours, the sum of every worker's recorded
  travel).
editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
library(dplyr)
library(readr)
library(ggplot2)
library(knitr)

tasks <- read_csv("../../data/all_tasks_clean.csv", show_col_types = FALSE)

fmt <- function(x, digits = 1) format(round(x, digits), nsmall = digits, big.mark = ",", trim = TRUE)

n_jobs_all <- n_distinct(tasks$job_id)
n_entries <- nrow(tasks)
n_workers <- n_distinct(tasks$gulper_id)
dates <- tasks$date
total_labour_h <- sum(tasks$duration_seconds) / 3600

job_labour <- tasks |>
  group_by(job_id) |>
  summarise(job_labour_h = sum(duration_seconds) / 3600, .groups = "drop")

# Errands, excluded from the analysis but described in Methods
errands <- tasks |> filter(task == "Travel", subtask == "On errand")

# Disposal-site travel entries, with overlapping entries within a job merged
# into trips: an entry starts a new trip if it begins after every earlier
# entry of the job has ended.
min_trip_min <- 5

travel <- tasks |>
  filter(task == "Travel", subtask == "Travel to disposal site") |>
  mutate(s = as.numeric(entry_start_time), e = as.numeric(entry_end_time)) |>
  arrange(job_id, s) |>
  group_by(job_id) |>
  mutate(prev_end = lag(cummax(e), default = -Inf), trip = cumsum(s > prev_end)) |>
  ungroup()

trips_all <- travel |>
  group_by(job_id, trip) |>
  summarise(
    start = min(s), end = max(e),
    gulpers = n_distinct(gulper_id),
    entries = n(),
    labour_h = sum(duration_seconds) / 3600,
    .groups = "drop"
  ) |>
  mutate(clock_min = (end - start) / 60, keep = clock_min >= min_trip_min)

trips <- trips_all |> filter(keep)
short <- trips_all |> filter(!keep)

summarise_jobs <- function(tr) {
  tr |>
    group_by(job_id) |>
    summarise(
      date = as.Date(substr(first(job_id), 1, 10)),
      trips = n(),
      gulpers = max(gulpers),
      clock_h = sum(end - start) / 3600,
      labour_h = sum(labour_h),
      .groups = "drop"
    ) |>
    left_join(job_labour, by = "job_id") |>
    mutate(share = 100 * labour_h / job_labour_h)
}

jobs <- summarise_jobs(trips)
jobs_unfiltered <- summarise_jobs(trips_all)

n_travel_jobs <- nrow(jobs)
n_trips <- nrow(trips)
mean_clock_h <- mean(jobs$clock_h)
median_clock_h <- median(jobs$clock_h)
mean_labour_h <- mean(jobs$labour_h)
median_labour_h <- median(jobs$labour_h)
mean_trip_min <- mean(trips$clock_min)
mean_gulpers_trip <- mean(trips$gulpers)
ratio <- sum(jobs$labour_h) / sum(jobs$clock_h)
travel_share_total <- 100 * sum(jobs$labour_h) / total_labour_h
```

# Introduction

Time-driven activity-based costing (TDABC) allocates costs to activities
based on the time resources spend on them. Manual pit emptying in Mzuzu
involves not only emptying the pit but also transporting the extracted
sludge to a disposal site, an activity that takes the whole team away
from the site and consumes labour capacity that must be costed. This
short manuscript answers the question: **What is the average time spent
per job travelling to the disposal site, when a team travels to the
disposal site?**

# Data and methods

The dataset (`data/all_tasks_clean.csv`) contains `r fmt(n_entries, 0)`
task entries recorded for `r n_jobs_all` pit emptying jobs between
`r format(min(dates), "%d %B %Y")` and
`r format(max(dates), "%d %B %Y")`, performed by `r n_workers`
anonymised pit emptiers ("gulpers"). Each entry records one worker
performing one task with a start time, end time and duration in seconds.
The Travel task has two subtasks: *On errand*, for short in-job errands,
and *Travel to disposal site*, for transporting sludge away from the
site. Only the second is analysed here. The `r nrow(errands)` errand
entries (`r fmt(sum(errands$duration_seconds) / 3600)` person-hours in
`r n_distinct(errands$job_id)` jobs) are excluded. Travel from the
gulpers' base to the job site is not among the recorded tasks, so the
figures below cover disposal trips made during a job only.

Gulpers travel to the disposal site together, so each trip appears in
the data as several near-identical entries, one per worker. Within each
job, entries that overlap in time are merged into a single **trip**
running from the earliest start to the latest end. Travel time per job
is then reported in two ways:

- **Elapsed time (clock hours):** the summed duration of the job's
  trips. A trip made by four gulpers together counts once. This is the
  headline measure, since it answers how long the team was away.
- **Labour time (person-hours):** the sum of `duration_seconds` over all
  disposal-site entries, i.e. every worker's travel added together. This
  is the measure relevant for TDABC, where labour capacity is the
  resource being costed, and it matches the measure used in the
  companion manuscript on total job time.

The population is the set of jobs in which the team travelled to the
disposal site. Genuine trips last
`r fmt(min(trips$clock_min), 0)`–`r fmt(max(trips$clock_min), 0)`
minutes. A few disposal-site entries are far shorter and cannot
represent a trip: `r nrow(short)` merged intervals of under
`r min_trip_min` minutes
(`r paste(fmt(sort(short$clock_min * 60), 0), collapse = ", ")`
seconds), each recorded by one or two workers and including a 2.9-second
entry already flagged in the data review. These are treated as
accidental taps and dropped, which removes
`r fmt(sum(short$labour_h) * 60)` person-minutes. One job
(`r setdiff(jobs_unfiltered$job_id, jobs$job_id)`) has only such an
entry and therefore does not count as a job with a disposal trip. The
effect of keeping these entries is reported in the Results.

The file is the cleaned version of the raw app export, produced by
`data-review/clean.R` and documented in `data-review/README.md`:
accidental taps shorter than 2 seconds were dropped, one subtask label
was harmonised and job identities were corrected from field knowledge.
None of the corrections touches a job with a disposal trip. Recorded
durations agree with the difference between end and start times to
within ±1 second, and no trip crosses midnight.

# Results

A team travelled to the disposal site in **`r n_travel_jobs` of the
`r n_jobs_all` observed jobs**, making `r n_trips` trips in total. In
those jobs, the team spent on average **`r fmt(mean_clock_h)` clock
hours (`r fmt(mean_clock_h * 60, 0)` minutes) per job travelling to the
disposal site** (median `r fmt(median_clock_h)` hours, range
`r fmt(min(jobs$clock_h))`–`r fmt(max(jobs$clock_h))`). In labour time
this is `r fmt(mean_labour_h)` person-hours per job (median
`r fmt(median_labour_h)`, range
`r fmt(min(jobs$labour_h))`–`r fmt(max(jobs$labour_h))`).

A job with a disposal trip had `r fmt(mean(jobs$trips))` trips on
average, each lasting `r fmt(mean_trip_min, 0)` minutes and made by
`r fmt(mean_gulpers_trip)` gulpers together, so labour time is
`r fmt(ratio)` times elapsed time. Within these jobs, disposal travel
took up `r fmt(mean(jobs$share), 0)` % of the job's labour time on
average (range `r fmt(min(jobs$share), 0)`–`r fmt(max(jobs$share), 0)`
%). Across all `r n_jobs_all` jobs, disposal travel accounts for
`r fmt(travel_share_total)` % of the `r fmt(total_labour_h)`
person-hours recorded. @tbl-jobs lists each job and @fig-jobs shows the
elapsed travel time per job.

```{r}
#| label: tbl-jobs
#| tbl-cap: "Disposal-site travel for each job in which a trip was made. Elapsed time counts a trip once regardless of how many gulpers made it; labour time adds every gulper's travel. Share is labour travel time as a percentage of the job's total labour time."
summary_row <- function(label, f) {
  tibble(
    `Job Date` = label,
    Trips = fmt(f(jobs$trips)), Gulpers = fmt(f(jobs$gulpers)),
    `Elapsed (h)` = f(jobs$clock_h), `Labour (person-h)` = f(jobs$labour_h),
    `Share of job (%)` = f(jobs$share)
  )
}

jobs |>
  transmute(
    `Job Date` = format(date, "%Y-%m-%d"),
    Trips = as.character(trips),
    Gulpers = as.character(gulpers),
    `Elapsed (h)` = clock_h,
    `Labour (person-h)` = labour_h,
    `Share of job (%)` = share
  ) |>
  bind_rows(summary_row("Mean", mean), summary_row("Median", median)) |>
  kable(digits = 1, align = "lrrrrr")
```

```{r}
#| label: fig-jobs
#| fig-cap: "Elapsed time spent travelling to the disposal site per job, for the jobs in which a trip was made, ordered by date. Labels give the number of trips. The dashed line is the mean."
#| fig-width: 6.5
#| fig-height: 3.2
jobs |>
  mutate(
    job_date = format(date, "%Y-%m-%d"),
    job_date = factor(job_date, levels = rev(job_date)),
    label = paste0(trips, ifelse(trips == 1, " trip", " trips"))
  ) |>
  ggplot(aes(x = clock_h, y = job_date)) +
  geom_col(fill = "#2a78d6", width = 0.7) +
  geom_text(aes(label = label), hjust = -0.15, size = 3, colour = "#52514e") +
  geom_vline(xintercept = mean_clock_h, linetype = "dashed", colour = "#52514e") +
  annotate("text", x = mean_clock_h, y = n_travel_jobs + 0.7,
           label = paste0("mean ", fmt(mean_clock_h), " h"),
           hjust = -0.05, size = 3, colour = "#52514e") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  coord_cartesian(clip = "off") +
  labs(x = "Elapsed travel time (clock hours)", y = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(colour = "#52514e"),
    plot.margin = margin(12, 8, 4, 4)
  )
```

Travel time per job is driven mainly by the number of trips: jobs with
one trip take about half an hour of elapsed time, and the two jobs with
the most travel
(`r paste(format(jobs$date[order(-jobs$clock_h)][1:2], "%d %B %Y"), collapse = " and ")`)
made
`r paste(sort(jobs$trips[order(-jobs$clock_h)][1:2], decreasing = TRUE), collapse = " and ")`
trips respectively.

As a sensitivity check, keeping the short entries dropped in the Methods
adds one job and changes the mean elapsed travel time to
`r fmt(mean(jobs_unfiltered$clock_h))` hours over
`r nrow(jobs_unfiltered)` jobs (`r fmt(mean(jobs_unfiltered$labour_h))`
person-hours); this is lower only because the added job contributes a
`r fmt(min(jobs_unfiltered$clock_h) * 60, 1)`-minute entry.

# Conclusion

When a pit emptying team in Mzuzu travels to the disposal site, it
spends on average **`r fmt(mean_clock_h)` clock hours
(`r fmt(mean_clock_h * 60, 0)` minutes) per job** on that travel,
corresponding to **`r fmt(mean_labour_h)` person-hours** of labour
because `r fmt(mean_gulpers_trip)` gulpers travel together. A single
trip takes about `r fmt(mean_trip_min, 0)` minutes, and jobs vary mainly
in how many trips they need. Such trips occurred in `r n_travel_jobs` of
the `r n_jobs_all` observed jobs. For TDABC purposes the person-hour
figure is the quantity to cost, and it should be applied only to jobs
that require a disposal trip.

# Appendix A: Relevant literature {.appendix}

The five references in the author's Zotero library most relevant to this
analysis, with one sentence on why each matters here:

1.  @yesaya2021cost apply time-driven activity-based costing to compost
    production in Blantyre, Malawi, making it the closest methodological
    precedent for turning the per-job travel times measured here into an
    activity cost.
2.  @wilcox2026formal analyse the costs of 23 formal manual emptying
    providers in Malawi and Uganda and find labour is what makes manual
    emptying expensive, which is why the person-hours that a team spends
    travelling to the disposal site are a key costing input.
3.  @chipeta2017designing designed and tested the pedal-powered Gulper
    pump in Mzuzu and estimated a pit can be emptied in 1–2 hours when
    trash content is low, so a half-hour disposal trip is a substantial
    addition to a job.
4.  @chiposa2017characterization characterise pit latrines in peri-urban
    Mzuzu, reporting that 90 % of households also dispose of domestic
    waste in their pits, which affects how much material must be removed
    and hence how many disposal trips a job needs.
5.  @burt2019costs estimate the cost of delivering pit emptying services
    in Kigali, Rwanda, at different scales, giving a comparison point
    for costs derived from the travel times reported here.

# References

::: {#refs}
:::
