---
title: "Time spent on manual pit emptying jobs in Mzuzu, Malawi: total job time and travel to the disposal site"
author:
  - name: Padraic Casserly
    email: pcasserly@ethz.ch
    affiliations:
      - name: Global Health Engineering, ETH Zürich
date: 2026-09-29
bibliography: references.bib
abstract: |
  We use task-level timing data recorded with the Gulper Timer app during
  ridealong observations of manual pit latrine emptying in Mzuzu, Malawi, to
  answer two questions that time-driven activity-based costing (TDABC) needs
  answered first: how much time is spent on a pit emptying job in total, and
  how much of a job is spent travelling to the sludge disposal site when the
  team makes such a trip. Time is reported in two ways throughout: labour
  time (person-hours, every worker's recorded time added together) and
  elapsed time (clock hours, with simultaneous work by several workers
  counted once).
editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
library(dplyr)
library(readr)
library(ggplot2)
library(knitr)

tasks <- read_csv("../../data/all_tasks_clean.csv", show_col_types = FALSE)

fmt <- function(x, digits = 1) format(round(x, digits), nsmall = digits, big.mark = ",", trim = TRUE)

# ---- Jobs: labour time and elapsed time -------------------------------------
# Elapsed time is the span from first start to last end on each recorded day;
# one job was recorded over two days, so its daily spans are added.
day_spans <- tasks |>
  group_by(job_id, date) |>
  summarise(span_h = as.numeric(max(entry_end_time) - min(entry_start_time)) / 3600,
            .groups = "drop")

jobs <- tasks |>
  group_by(job_id) |>
  summarise(
    date = min(date),
    n_days = n_distinct(date),
    n_workers = n_distinct(gulper_id),
    n_entries = n(),
    labour_h = sum(duration_seconds) / 3600,
    .groups = "drop"
  ) |>
  left_join(day_spans |> group_by(job_id) |> summarise(elapsed_h = sum(span_h), .groups = "drop"),
            by = "job_id") |>
  arrange(date, job_id) |>
  group_by(date) |>
  mutate(job_label = if (n() > 1) paste0(format(date, "%Y-%m-%d"), " (", row_number(), ")")
                     else format(date, "%Y-%m-%d")) |>
  ungroup()

total_labour_h <- sum(jobs$labour_h)
total_elapsed_h <- sum(jobs$elapsed_h)
n_jobs <- nrow(jobs)
n_workers <- n_distinct(tasks$gulper_id)
n_entries <- nrow(tasks)
rest_h <- sum(tasks$duration_seconds[tasks$task == "Rest"]) / 3600

# Sensitivity: the entries flagged timer_suspect capped at 50 minutes
capped_labour_h <- tasks |>
  mutate(d = if_else(grepl("timer_suspect", flags), pmin(duration_seconds, 3000), duration_seconds)) |>
  summarise(h = sum(d) / 3600) |>
  pull(h)
n_suspect <- sum(grepl("timer_suspect", tasks$flags))

# ---- Travel to the disposal site ------------------------------------------
errands <- tasks |> filter(task == "Travel", subtask == "On errand")
min_trip_min <- 5

travel <- tasks |>
  filter(task == "Travel", subtask == "Travel to disposal site") |>
  mutate(s = as.numeric(entry_start_time), e = as.numeric(entry_end_time)) |>
  arrange(job_id, s) |>
  group_by(job_id) |>
  mutate(prev_end = lag(cummax(e), default = -Inf), trip = cumsum(s > prev_end)) |>
  ungroup()

trips_all <- travel |>
  group_by(job_id, trip) |>
  summarise(start = min(s), end = max(e), gulpers = n_distinct(gulper_id),
            entries = n(), labour_h = sum(duration_seconds) / 3600, .groups = "drop") |>
  mutate(clock_min = (end - start) / 60, keep = clock_min >= min_trip_min)
trips <- trips_all |> filter(keep)
short <- trips_all |> filter(!keep)

summarise_travel <- function(tr) {
  tr |>
    group_by(job_id) |>
    summarise(trips = n(), gulpers = max(gulpers),
              travel_clock_h = sum(end - start) / 3600,
              travel_labour_h = sum(labour_h), .groups = "drop") |>
    left_join(jobs |> select(job_id, date, job_label, labour_h), by = "job_id") |>
    mutate(share = 100 * travel_labour_h / labour_h) |>
    arrange(date, job_id)
}
travel_jobs <- summarise_travel(trips)
travel_jobs_unfiltered <- summarise_travel(trips_all)

n_travel_jobs <- nrow(travel_jobs)
n_trips <- nrow(trips)
mean_clock_h <- mean(travel_jobs$travel_clock_h)
median_clock_h <- median(travel_jobs$travel_clock_h)
mean_travel_labour_h <- mean(travel_jobs$travel_labour_h)
median_travel_labour_h <- median(travel_jobs$travel_labour_h)
mean_trip_min <- mean(trips$clock_min)
mean_gulpers_trip <- mean(trips$gulpers)
ratio <- sum(travel_jobs$travel_labour_h) / sum(travel_jobs$travel_clock_h)
travel_share_total <- 100 * sum(travel_jobs$travel_labour_h) / total_labour_h

# Shared plot styling
bar_theme <- function() {
  theme_minimal(base_size = 10) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text = element_text(colour = "#52514e"),
      plot.margin = margin(12, 8, 4, 4)
    )
}
```

# Introduction

Time-driven activity-based costing (TDABC) allocates costs to activities
based on the time resources spend on them. A first step is therefore to
establish how much time the service being costed consumes, and how that
time is split between activities that every job needs and activities
that only some jobs need. Manual pit emptying in Mzuzu involves not only
emptying the pit but, for some jobs, transporting the extracted sludge
to a disposal site, an activity that takes the whole team away from the
site. This manuscript answers two questions:

1. **What is the total amount of time spent on pit emptying jobs?**
2. **What is the average time spent per job travelling to the disposal
   site, when a team travels to the disposal site?**

# Data and methods

## Data

The dataset (`data/all_tasks_clean.csv`) contains `r fmt(n_entries, 0)`
task entries recorded for `r n_jobs` pit emptying jobs between
`r format(min(jobs$date), "%d %B %Y")` and
`r format(max(jobs$date), "%d %B %Y")`, performed by `r n_workers`
anonymised pit emptiers ("gulpers"). Each entry records one worker
performing one task (e.g. Fluidizing, Extraction, Rest) with a start
time, end time and duration in seconds.

The file is the cleaned version of the raw app export, produced by
`data-review/clean.R` and documented in `data-review/README.md`.
Relative to the raw export, six accidental taps shorter than 2 seconds
were dropped, the Equipment subtask "Repair" was recoded to "Repairing",
and job identities were corrected from field knowledge: the recordings
filed as 2026-06-12_01 and 2026-06-13_01 are one job carried out over
two days by two crews, and the two recording sessions filed under
2026-06-14_02 are two separate jobs. The file has no missing values
except `subtask`, which is empty by design for the Rest and Disposal
tasks, and recorded durations agree with the difference between end and
start times to within ±1 second. Sixteen entries longer than 50 minutes
are kept as recorded; `r n_suspect` of them, in one job, look like
timers left running over a break and are used in a sensitivity check.
Jobs are identified below by their date; where two jobs were recorded on
the same day they are numbered (1) and (2).

## Two measures of time

Several gulpers usually work simultaneously on the same job, so "time
spent" is measured in two ways throughout:

- **Labour time (person-hours):** the sum of `duration_seconds` over all
  entries, i.e. every worker's recorded time added together. This is the
  measure relevant for TDABC, where labour capacity is the resource
  being costed.
- **Elapsed time (clock hours):** for each job, the interval from the
  earliest `entry_start_time` to the latest `entry_end_time` on each
  recorded day, summed across days and jobs. This measures how long the
  jobs took on the calendar.

All recorded tasks, including Rest, Communication and in-job Travel, are
counted as time spent on the job. Travel from the gulpers' base to the
job site is not among the recorded tasks.

## Travel to the disposal site

The Travel task has two subtasks: *On errand*, for short in-job errands,
and *Travel to disposal site*, for transporting sludge away from the
site. Only the second is analysed for question 2; the `r nrow(errands)`
errand entries (`r fmt(sum(errands$duration_seconds) / 3600)`
person-hours in `r n_distinct(errands$job_id)` jobs) remain part of
total job time but are not counted as disposal travel.

Gulpers travel to the disposal site together, so each trip appears in
the data as several near-identical entries, one per worker. Within each
job, entries that overlap in time are merged into a single **trip**
running from the earliest start to the latest end. A job's elapsed
travel time is the summed duration of its trips, counting a trip made by
four gulpers once; its labour travel time adds every gulper's entry.

The population for question 2 is the set of jobs in which the team
travelled to the disposal site. Genuine trips last
`r fmt(min(trips$clock_min), 0)`–`r fmt(max(trips$clock_min), 0)`
minutes. A few disposal-site entries are far shorter and cannot
represent a trip: `r nrow(short)` merged intervals of under
`r min_trip_min` minutes
(`r paste(fmt(sort(short$clock_min * 60), 0), collapse = ", ")`
seconds), each recorded by one or two workers. These are treated as
accidental taps and dropped, which removes
`r fmt(sum(short$labour_h) * 60)` person-minutes. One job has only such
an entry and therefore does not count as a job with a disposal trip.

# Results

## Total time spent on pit emptying jobs

Across the `r n_jobs` observed jobs, gulpers spent a total of
**`r fmt(total_labour_h)` person-hours** on pit emptying. The jobs took
a total of **`r fmt(total_elapsed_h)` elapsed clock hours**, so on
average `r fmt(total_labour_h / total_elapsed_h)` gulpers' worth of time
was being recorded at any moment of a job. Per job, the mean labour time
was `r fmt(mean(jobs$labour_h))` person-hours (median
`r fmt(median(jobs$labour_h))`, range
`r fmt(min(jobs$labour_h))`–`r fmt(max(jobs$labour_h))`) and the mean
elapsed time was `r fmt(mean(jobs$elapsed_h))` hours (median
`r fmt(median(jobs$elapsed_h))`, range
`r fmt(min(jobs$elapsed_h))`–`r fmt(max(jobs$elapsed_h))`).

@tbl-task breaks the labour time down by main task. Rest is the largest
single category (`r fmt(rest_h)` person-hours,
`r fmt(100 * rest_h / total_labour_h)` % of the total).

```{r}
#| label: tbl-task
#| tbl-cap: "Labour time by main task, summed over all 36 jobs."
tasks |>
  group_by(Task = task) |>
  summarise(Entries = n(), `Person-hours` = sum(duration_seconds) / 3600, .groups = "drop") |>
  arrange(desc(`Person-hours`)) |>
  mutate(`Share (%)` = 100 * `Person-hours` / sum(`Person-hours`)) |>
  bind_rows(tibble(Task = "Total", Entries = n_entries,
                   `Person-hours` = total_labour_h, `Share (%)` = 100)) |>
  kable(digits = 1, format.args = list(big.mark = ","))
```

@fig-task shows the same breakdown, with tasks coloured by the kind of
work they belong to.

```{r}
#| label: fig-task
#| fig-cap: "Labour time by main task across all 36 jobs, coloured by activity group. Labels give person-hours and share of total labour time."
#| fig-width: 6.5
#| fig-height: 4.2
task_groups <- c(
  Fluidizing        = "Emptying the pit",
  Extraction        = "Emptying the pit",
  `Sludge handling` = "Emptying the pit",
  Disposal          = "Emptying the pit",
  Earthworks        = "Site and structure works",
  Superstructure    = "Site and structure works",
  Equipment         = "Support tasks",
  Communication     = "Support tasks",
  Travel            = "Support tasks",
  Rest              = "Rest"
)
group_colours <- c(
  "Emptying the pit"         = "#2a78d6",
  "Site and structure works" = "#eb6834",
  "Support tasks"            = "#1baf7a",
  "Rest"                     = "#4a3aa7"
)

task_summary <- tasks |>
  group_by(task) |>
  summarise(hours = sum(duration_seconds) / 3600, .groups = "drop") |>
  mutate(
    share = 100 * hours / sum(hours),
    group = factor(task_groups[task], levels = names(group_colours)),
    task = reorder(task, hours),
    label = paste0(fmt(hours), " h (", fmt(share), " %)")
  )

ggplot(task_summary, aes(x = hours, y = task, fill = group)) +
  geom_col(width = 0.72) +
  geom_text(aes(label = label), hjust = -0.08, size = 3, colour = "#52514e") +
  scale_fill_manual(values = group_colours, name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.28))) +
  labs(x = "Labour time (person-hours)", y = NULL) +
  bar_theme() +
  theme(
    legend.position = "top",
    legend.justification = "left",
    legend.key.size = unit(0.4, "cm"),
    axis.text.y = element_text(colour = "#0b0b0b", size = 10)
  )

group_hours <- tapply(task_summary$hours, task_summary$group, sum)
```

Grouped this way, the tasks that directly empty the pit (fluidizing,
extraction, sludge handling and disposal) account for
`r fmt(group_hours[["Emptying the pit"]])` person-hours
(`r fmt(100 * group_hours[["Emptying the pit"]] / total_labour_h)` %),
about the same as Rest (`r fmt(group_hours[["Rest"]])` person-hours).
Support tasks take `r fmt(group_hours[["Support tasks"]])` person-hours
(`r fmt(100 * group_hours[["Support tasks"]] / total_labour_h)` %) and
site and structure works
`r fmt(group_hours[["Site and structure works"]])` person-hours
(`r fmt(100 * group_hours[["Site and structure works"]] / total_labour_h)`
%).

@fig-jobs shows the labour time of each job. Labour time varies widely
between jobs; @tbl-jobs in the appendix lists the values for every job.

```{r}
#| label: fig-jobs
#| fig-cap: "Labour time (person-hours) per pit emptying job, ordered by date. The dashed line is the mean."
#| fig-width: 6.5
#| fig-height: 6
jobs |>
  mutate(job_label = factor(job_label, levels = rev(job_label))) |>
  ggplot(aes(x = labour_h, y = job_label)) +
  geom_col(fill = "#2a78d6", width = 0.7) +
  geom_vline(xintercept = mean(jobs$labour_h), linetype = "dashed", colour = "#52514e") +
  annotate("text", x = mean(jobs$labour_h), y = n_jobs + 0.6,
           label = paste0("mean ", fmt(mean(jobs$labour_h)), " h"),
           hjust = -0.05, size = 3, colour = "#52514e") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  coord_cartesian(clip = "off") +
  labs(x = "Labour time (person-hours)", y = NULL) +
  bar_theme()
```

## Time spent travelling to the disposal site

A team travelled to the disposal site in **`r n_travel_jobs` of the
`r n_jobs` observed jobs**, making `r n_trips` trips in total. In those
jobs, the team spent on average **`r fmt(mean_clock_h)` clock hours
(`r fmt(mean_clock_h * 60, 0)` minutes) per job travelling to the
disposal site** (median `r fmt(median_clock_h)` hours, range
`r fmt(min(travel_jobs$travel_clock_h))`–`r fmt(max(travel_jobs$travel_clock_h))`).
In labour time this is `r fmt(mean_travel_labour_h)` person-hours per
job (median `r fmt(median_travel_labour_h)`, range
`r fmt(min(travel_jobs$travel_labour_h))`–`r fmt(max(travel_jobs$travel_labour_h))`).

A job with a disposal trip had `r fmt(mean(travel_jobs$trips))` trips
on average, each lasting `r fmt(mean_trip_min, 0)` minutes and made by
`r fmt(mean_gulpers_trip)` gulpers together, so labour time is
`r fmt(ratio)` times elapsed time. Within these jobs, disposal travel
took up `r fmt(mean(travel_jobs$share), 0)` % of the job's labour time
on average (range
`r fmt(min(travel_jobs$share), 0)`–`r fmt(max(travel_jobs$share), 0)`
%). Across all `r n_jobs` jobs, disposal travel accounts for
`r fmt(travel_share_total)` % of the `r fmt(total_labour_h)`
person-hours recorded. @tbl-travel lists each job with a trip and
@fig-travel shows the elapsed travel time per job.

```{r}
#| label: tbl-travel
#| tbl-cap: "Disposal-site travel for each job in which a trip was made. Elapsed time counts a trip once regardless of how many gulpers made it; labour time adds every gulper's travel. Share is labour travel time as a percentage of the job's total labour time."
summary_row <- function(label, f) {
  tibble(
    `Job Date` = label,
    Trips = fmt(f(travel_jobs$trips)), Gulpers = fmt(f(travel_jobs$gulpers)),
    `Elapsed (h)` = f(travel_jobs$travel_clock_h),
    `Labour (person-h)` = f(travel_jobs$travel_labour_h),
    `Share of job (%)` = f(travel_jobs$share)
  )
}

travel_jobs |>
  transmute(
    `Job Date` = job_label,
    Trips = as.character(trips),
    Gulpers = as.character(gulpers),
    `Elapsed (h)` = travel_clock_h,
    `Labour (person-h)` = travel_labour_h,
    `Share of job (%)` = share
  ) |>
  bind_rows(summary_row("Mean", mean), summary_row("Median", median)) |>
  kable(digits = 1, align = "lrrrrr")
```

```{r}
#| label: fig-travel
#| fig-cap: "Elapsed time spent travelling to the disposal site per job, for the jobs in which a trip was made, ordered by date. Labels give the number of trips. The dashed line is the mean."
#| fig-width: 6.5
#| fig-height: 3.2
travel_jobs |>
  mutate(
    job_label = factor(job_label, levels = rev(job_label)),
    label = paste0(trips, ifelse(trips == 1, " trip", " trips"))
  ) |>
  ggplot(aes(x = travel_clock_h, y = job_label)) +
  geom_col(fill = "#2a78d6", width = 0.7) +
  geom_text(aes(label = label), hjust = -0.15, size = 3, colour = "#52514e") +
  geom_vline(xintercept = mean_clock_h, linetype = "dashed", colour = "#52514e") +
  annotate("text", x = mean_clock_h, y = n_travel_jobs + 0.7,
           label = paste0("mean ", fmt(mean_clock_h), " h"),
           hjust = -0.05, size = 3, colour = "#52514e") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  coord_cartesian(clip = "off") +
  labs(x = "Elapsed travel time (clock hours)", y = NULL) +
  bar_theme()
```

Travel time per job is driven mainly by the number of trips: jobs with
one trip take about half an hour of elapsed time, and the two jobs with
the most travel
(`r paste(travel_jobs$job_label[order(-travel_jobs$travel_clock_h)][1:2], collapse = " and ")`)
made
`r paste(sort(travel_jobs$trips[order(-travel_jobs$travel_clock_h)][1:2], decreasing = TRUE), collapse = " and ")`
trips respectively.

## Sensitivity checks

Capping the `r n_suspect` entries that look like timers left running at
50 minutes would lower the total labour time to
`r fmt(capped_labour_h)` person-hours, a change of
`r fmt(total_labour_h - capped_labour_h)` person-hours
(`r fmt(100 * (total_labour_h - capped_labour_h) / total_labour_h)` %).
Keeping the short disposal-site entries dropped in the Methods adds one
job and changes the mean elapsed travel time to
`r fmt(mean(travel_jobs_unfiltered$travel_clock_h))` hours over
`r nrow(travel_jobs_unfiltered)` jobs
(`r fmt(mean(travel_jobs_unfiltered$travel_labour_h))` person-hours);
this is lower only because the added job contributes a
`r fmt(min(travel_jobs_unfiltered$travel_clock_h) * 60, 1)`-minute
entry.

# Conclusion

The `r n_jobs` observed pit emptying jobs consumed
**`r fmt(total_labour_h)` person-hours** of recorded labour
(`r fmt(total_elapsed_h)` elapsed clock hours), a mean of
`r fmt(mean(jobs$labour_h))` person-hours per job. When a team travels
to the disposal site, which happened in `r n_travel_jobs` of the
`r n_jobs` jobs, it spends on average **`r fmt(mean_clock_h)` clock
hours (`r fmt(mean_clock_h * 60, 0)` minutes) per job** on that travel,
corresponding to **`r fmt(mean_travel_labour_h)` person-hours** of
labour because `r fmt(mean_gulpers_trip)` gulpers travel together. A
single trip takes about `r fmt(mean_trip_min, 0)` minutes, and jobs vary
mainly in how many trips they need. For TDABC purposes the person-hour
figures are the quantities to cost: total labour time applies to every
job, and disposal travel should be added only to the jobs that require a
trip.

# Appendix A: Job-level results {.appendix}

```{r}
#| label: tbl-jobs
#| tbl-cap: "Workers, labour time, elapsed time and disposal-site travel for each job. Travel columns are blank for jobs without a trip."
options(knitr.kable.NA = "")
jobs |>
  left_join(travel_jobs |> select(job_id, travel_clock_h, travel_labour_h), by = "job_id") |>
  transmute(
    `Job Date` = job_label,
    Gulpers = n_workers,
    Entries = n_entries,
    `Labour (person-h)` = labour_h,
    `Elapsed (h)` = elapsed_h,
    `Travel elapsed (h)` = travel_clock_h,
    `Travel labour (person-h)` = travel_labour_h
  ) |>
  bind_rows(tibble(
    `Job Date` = "Total", Gulpers = n_workers, Entries = n_entries,
    `Labour (person-h)` = total_labour_h, `Elapsed (h)` = total_elapsed_h,
    `Travel elapsed (h)` = sum(travel_jobs$travel_clock_h),
    `Travel labour (person-h)` = sum(travel_jobs$travel_labour_h)
  )) |>
  kable(digits = 1, format.args = list(big.mark = ","))
```

# Appendix B: Relevant literature {.appendix}

The five references in the author's Zotero library most relevant to this
dataset, with one sentence on why each matters here:

1.  @yesaya2021cost apply time-driven activity-based costing to compost
    production in Blantyre, Malawi, making it the closest methodological
    precedent for turning task-level timing data like these into
    activity costs.
2.  @wilcox2026formal analyse the costs of 23 formal manual emptying
    providers in Malawi and Uganda and find labour is what makes manual
    emptying expensive, which is why the person-hours measured here, for
    the whole job and for disposal travel, are the key costing inputs.
3.  @chipeta2017designing designed and tested the pedal-powered Gulper
    pump in Mzuzu and estimated that a pit can be emptied in 1–2 hours
    when trash content is low, a benchmark for the job durations recorded
    here and one against which a half-hour disposal trip is substantial.
4.  @chiposa2017characterization characterise pit latrines in peri-urban
    Mzuzu, reporting that 90 % of households also dispose of domestic
    waste in their pits, which helps explain the time spent on tasks such
    as pulling out trash and affects how much material must be removed.
5.  @burt2019costs estimate the cost of delivering pit emptying services
    in Kigali, Rwanda, at different scales, giving a comparison point for
    costs derived from the labour and travel times reported here.

# References

::: {#refs}
:::
