---
title: "Total time spent on pit emptying jobs in Mzuzu, Malawi"
author:
  - name: Padraic Casserly
    email: pcasserly@ethz.ch
    affiliations:
      - name: Global Health Engineering, ETH Zürich
date: 2026-09-29
abstract: |
  We use task-level timing data recorded with the Gulper Timer app during
  ridealong observations of manual pit latrine emptying in Mzuzu, Malawi, to
  quantify the total time spent on pit emptying jobs. Time is reported in two
  ways: labour time (person-hours, the sum of all task durations across all
  workers) and elapsed time (clock hours from the first to the last recorded
  task of each job).
---

```{r}
#| label: setup
library(dplyr)
library(readr)
library(ggplot2)
library(knitr)

tasks <- read_csv("../../data/all_tasks.csv", show_col_types = FALSE)

fmt <- function(x, digits = 1) format(round(x, digits), nsmall = digits, big.mark = ",")

jobs <- tasks |>
  group_by(job_id) |>
  summarise(
    date = as.Date(substr(first(job_id), 1, 10)),
    n_workers = n_distinct(gulper_id),
    n_entries = n(),
    labour_h = sum(duration_seconds) / 3600,
    elapsed_h = as.numeric(max(entry_end_time) - min(entry_start_time)) / 3600,
    .groups = "drop"
  )

total_labour_h <- sum(jobs$labour_h)
total_elapsed_h <- sum(jobs$elapsed_h)
n_jobs <- nrow(jobs)
n_workers <- n_distinct(tasks$gulper_id)
n_entries <- nrow(tasks)

rest_h <- sum(tasks$duration_seconds[tasks$task == "Rest"]) / 3600
```

# Introduction

Time-driven activity-based costing (TDABC) allocates costs to activities based
on the time resources spend on them. A first step is therefore to establish how
much time is consumed by the service being costed. This short manuscript
answers the question: **What is the total amount of time spent on pit emptying
jobs?**

# Data and methods

The dataset (`data/all_tasks.csv`) contains `r fmt(n_entries, 0)` task entries
recorded for `r n_jobs` pit emptying jobs between
`r format(min(jobs$date), "%d %B %Y")` and `r format(max(jobs$date), "%d %B %Y")`,
performed by `r n_workers` anonymised pit emptiers ("gulpers"). Each entry records
one worker performing one task (e.g. Fluidizing, Extraction, Rest) with a start
time, end time and duration in seconds.

Because several gulpers often work simultaneously on the same job, "time spent"
can be measured in two ways:

- **Labour time (person-hours):** the sum of `duration_seconds` over all entries,
  i.e. every worker's recorded time added together. This is the measure relevant
  for TDABC, where labour capacity is the resource being costed.
- **Elapsed time (clock hours):** for each job, the interval from the earliest
  `entry_start_time` to the latest `entry_end_time`, summed across jobs. This
  measures how long the jobs took on the calendar.

All recorded tasks, including Rest, Communication and in-job Travel (errands and
trips to the disposal site), are counted as time spent on the job. Travel from
the gulpers' base to the job site is not part of the task categories recorded.

Data checks: the file has no missing values except `subtask`, which is empty by
design for the Rest and Disposal tasks. Recorded durations agree with the
difference between end and start times to within ±1 second (rounding). No job
crosses midnight, so elapsed time can be computed from clock times directly.

# Results

Across the `r n_jobs` observed jobs, gulpers spent a total of
**`r fmt(total_labour_h)` person-hours** on pit emptying. The jobs took a total of
**`r fmt(total_elapsed_h)` elapsed clock hours**, so on average
`r fmt(total_labour_h / total_elapsed_h)` gulpers' worth of time was being recorded
at any moment of a job. Per job, the mean labour time was
`r fmt(mean(jobs$labour_h))` person-hours (median `r fmt(median(jobs$labour_h))`,
range `r fmt(min(jobs$labour_h))`–`r fmt(max(jobs$labour_h))`) and the mean
elapsed time was `r fmt(mean(jobs$elapsed_h))` hours (median
`r fmt(median(jobs$elapsed_h))`, range
`r fmt(min(jobs$elapsed_h))`–`r fmt(max(jobs$elapsed_h))`).

@tbl-task breaks the labour time down by main task. Rest is the largest single
category (`r fmt(rest_h)` person-hours, `r fmt(100 * rest_h / total_labour_h)` %
of the total).

```{r}
#| label: tbl-task
#| tbl-cap: "Labour time by main task, summed over all 36 jobs."
tasks |>
  group_by(Task = task) |>
  summarise(
    Entries = n(),
    `Person-hours` = sum(duration_seconds) / 3600,
    .groups = "drop"
  ) |>
  arrange(desc(`Person-hours`)) |>
  mutate(`Share (%)` = 100 * `Person-hours` / sum(`Person-hours`)) |>
  bind_rows(tibble(
    Task = "Total", Entries = n_entries,
    `Person-hours` = total_labour_h, `Share (%)` = 100
  )) |>
  kable(digits = 1, format.args = list(big.mark = ","))
```

@fig-jobs shows the labour time of each job. Labour time varies widely between jobs;
@tbl-jobs in the appendix lists the values for every job.

```{r}
#| label: fig-jobs
#| fig-cap: "Labour time (person-hours) per pit emptying job, ordered by date."
#| fig-width: 6.5
#| fig-height: 6
jobs |>
  mutate(job_id = factor(job_id, levels = rev(job_id))) |>
  ggplot(aes(x = labour_h, y = job_id)) +
  geom_col(fill = "#2a78d6", width = 0.7) +
  geom_vline(xintercept = mean(jobs$labour_h), linetype = "dashed", colour = "#52514e") +
  annotate("text", x = mean(jobs$labour_h), y = n_jobs + 0.6,
           label = paste0("mean ", fmt(mean(jobs$labour_h)), " h"),
           hjust = -0.05, size = 3, colour = "#52514e") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  coord_cartesian(clip = "off") +
  labs(x = "Labour time (person-hours)", y = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(colour = "#52514e"),
    plot.margin = margin(12, 8, 4, 4)
  )
```

# Conclusion

The total amount of time spent on the `r n_jobs` observed pit emptying jobs was
**`r fmt(total_labour_h)` person-hours** of recorded labour
(`r fmt(total_elapsed_h)` elapsed clock hours). For TDABC purposes the
person-hour figure is the relevant quantity, since it reflects the labour
capacity consumed by the service.

# Appendix {.appendix}

```{r}
#| label: tbl-jobs
#| tbl-cap: "Workers, labour time and elapsed time for each job."
jobs |>
  transmute(
    Job = job_id,
    Gulpers = n_workers,
    Entries = n_entries,
    `Labour (person-h)` = labour_h,
    `Elapsed (h)` = elapsed_h
  ) |>
  bind_rows(tibble(
    Job = "Total", Gulpers = n_workers, Entries = n_entries,
    `Labour (person-h)` = total_labour_h, `Elapsed (h)` = total_elapsed_h
  )) |>
  kable(digits = 1, format.args = list(big.mark = ","))
```
