---
title: "Total time spent on pit emptying jobs in Mzuzu, Malawi"
author:
  - name: Padraic Casserly
    email: pcasserly@ethz.ch
    affiliations:
      - name: Global Health Engineering, ETH Zürich
date: 2026-09-29
bibliography: references.bib
abstract: |
  We use task-level timing data recorded with the Gulper Timer app during
  ridealong observations of manual pit latrine emptying in Mzuzu, Malawi, to
  quantify the total time spent on pit emptying jobs. Time is reported in two
  ways: labour time (person-hours, the sum of all task durations across all
  workers) and elapsed time (clock hours from the first to the last recorded
  task of each job).
---

```{r}
#| label: setup
library(dplyr)
library(readr)
library(ggplot2)
library(knitr)

tasks <- read_csv("../../data/all_tasks_clean.csv", show_col_types = FALSE)

fmt <- function(x, digits = 1) format(round(x, digits), nsmall = digits, big.mark = ",", trim = TRUE)

# Elapsed time is the span from first start to last end on each recorded
# day; one job was recorded over two days, so its daily spans are added.
day_spans <- tasks |>
  group_by(job_id, date) |>
  summarise(span_h = as.numeric(max(entry_end_time) - min(entry_start_time)) / 3600,
            .groups = "drop")

jobs <- tasks |>
  group_by(job_id) |>
  summarise(
    date = min(date),
    n_days = n_distinct(date),
    n_workers = n_distinct(gulper_id),
    n_entries = n(),
    labour_h = sum(duration_seconds) / 3600,
    .groups = "drop"
  ) |>
  left_join(day_spans |> group_by(job_id) |> summarise(elapsed_h = sum(span_h), .groups = "drop"),
            by = "job_id")

# Sensitivity: the four entries flagged timer_suspect capped at 50 minutes
capped_labour_h <- tasks |>
  mutate(d = if_else(grepl("timer_suspect", flags), pmin(duration_seconds, 3000), duration_seconds)) |>
  summarise(h = sum(d) / 3600) |>
  pull(h)
n_suspect <- sum(grepl("timer_suspect", tasks$flags))

total_labour_h <- sum(jobs$labour_h)
total_elapsed_h <- sum(jobs$elapsed_h)
n_jobs <- nrow(jobs)
n_workers <- n_distinct(tasks$gulper_id)
n_entries <- nrow(tasks)

rest_h <- sum(tasks$duration_seconds[tasks$task == "Rest"]) / 3600
```

# Introduction

Time-driven activity-based costing (TDABC) allocates costs to activities based
on the time resources spend on them. A first step is therefore to establish how
much time is consumed by the service being costed. This short manuscript
answers the question: **What is the total amount of time spent on pit emptying
jobs?**

# Data and methods

The dataset (`data/all_tasks_clean.csv`) contains `r fmt(n_entries, 0)` task entries
recorded for `r n_jobs` pit emptying jobs between
`r format(min(jobs$date), "%d %B %Y")` and `r format(max(jobs$date), "%d %B %Y")`,
performed by `r n_workers` anonymised pit emptiers ("gulpers"). Each entry records
one worker performing one task (e.g. Fluidizing, Extraction, Rest) with a start
time, end time and duration in seconds.

Because several gulpers often work simultaneously on the same job, "time spent"
can be measured in two ways:

- **Labour time (person-hours):** the sum of `duration_seconds` over all entries,
  i.e. every worker's recorded time added together. This is the measure relevant
  for TDABC, where labour capacity is the resource being costed.
- **Elapsed time (clock hours):** for each job, the interval from the earliest
  `entry_start_time` to the latest `entry_end_time`, summed across jobs. For
  the one job recorded over two days the two daily intervals are added. This
  measures how long the jobs took on the calendar.

All recorded tasks, including Rest, Communication and in-job Travel (errands and
trips to the disposal site), are counted as time spent on the job. Travel from
the gulpers' base to the job site is not part of the task categories recorded.

The file is the cleaned version of the raw app export, produced by
`data-review/clean.R` and documented in `data-review/README.md`. Relative to
the raw export, six accidental taps shorter than 2 seconds were dropped, the
Equipment subtask "Repair" was recoded to "Repairing", and job identities were
corrected from field knowledge: the recordings filed as 2026-06-12_01 and
2026-06-13_01 are one job carried out over two days by two crews, and the two
recording sessions filed under 2026-06-14_02 are two separate jobs
(2026-06-14_02a and 2026-06-14_02b). The file has no missing values except
`subtask`, which is empty by design for the Rest and Disposal tasks, and
recorded durations agree with the difference between end and start times to
within ±1 second. Sixteen entries longer than 50 minutes are kept as recorded;
`r n_suspect` of them, in 2026-09-03_03, look like timers left running over a
break and are used in a sensitivity check below.

# Results

Across the `r n_jobs` observed jobs, gulpers spent a total of
**`r fmt(total_labour_h)` person-hours** on pit emptying. The jobs took a total of
**`r fmt(total_elapsed_h)` elapsed clock hours**, so on average
`r fmt(total_labour_h / total_elapsed_h)` gulpers' worth of time was being recorded
at any moment of a job. Per job, the mean labour time was
`r fmt(mean(jobs$labour_h))` person-hours (median `r fmt(median(jobs$labour_h))`,
range `r fmt(min(jobs$labour_h))`–`r fmt(max(jobs$labour_h))`) and the mean
elapsed time was `r fmt(mean(jobs$elapsed_h))` hours (median
`r fmt(median(jobs$elapsed_h))`, range
`r fmt(min(jobs$elapsed_h))`–`r fmt(max(jobs$elapsed_h))`). Capping the
`r n_suspect` suspect entries at 50 minutes would lower the total labour time
to `r fmt(capped_labour_h)` person-hours, a change of
`r fmt(total_labour_h - capped_labour_h)` person-hours
(`r fmt(100 * (total_labour_h - capped_labour_h) / total_labour_h)` %).

@tbl-task breaks the labour time down by main task. Rest is the largest single
category (`r fmt(rest_h)` person-hours, `r fmt(100 * rest_h / total_labour_h)` %
of the total).

```{r}
#| label: tbl-task
#| tbl-cap: "Labour time by main task, summed over all 36 jobs."
tasks |>
  group_by(Task = task) |>
  summarise(
    Entries = n(),
    `Person-hours` = sum(duration_seconds) / 3600,
    .groups = "drop"
  ) |>
  arrange(desc(`Person-hours`)) |>
  mutate(`Share (%)` = 100 * `Person-hours` / sum(`Person-hours`)) |>
  bind_rows(tibble(
    Task = "Total", Entries = n_entries,
    `Person-hours` = total_labour_h, `Share (%)` = 100
  )) |>
  kable(digits = 1, format.args = list(big.mark = ","))
```

@fig-task shows the same breakdown, with tasks coloured by the kind of work
they belong to.

```{r}
#| label: fig-task
#| fig-cap: "Labour time by main task across all 36 jobs, coloured by activity group. Labels give person-hours and share of total labour time."
#| fig-width: 6.5
#| fig-height: 4.2
task_groups <- c(
  Fluidizing        = "Emptying the pit",
  Extraction        = "Emptying the pit",
  `Sludge handling` = "Emptying the pit",
  Disposal          = "Emptying the pit",
  Earthworks        = "Site and structure works",
  Superstructure    = "Site and structure works",
  Equipment         = "Support tasks",
  Communication     = "Support tasks",
  Travel            = "Support tasks",
  Rest              = "Rest"
)
group_colours <- c(
  "Emptying the pit"         = "#2a78d6",
  "Site and structure works" = "#eb6834",
  "Support tasks"            = "#1baf7a",
  "Rest"                     = "#4a3aa7"
)

task_summary <- tasks |>
  group_by(task) |>
  summarise(hours = sum(duration_seconds) / 3600, .groups = "drop") |>
  mutate(
    share = 100 * hours / sum(hours),
    group = factor(task_groups[task], levels = names(group_colours)),
    task = reorder(task, hours),
    label = paste0(fmt(hours), " h (", fmt(share), " %)")
  )

ggplot(task_summary, aes(x = hours, y = task, fill = group)) +
  geom_col(width = 0.72) +
  geom_text(aes(label = label), hjust = -0.08, size = 3, colour = "#52514e") +
  scale_fill_manual(values = group_colours, name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.28))) +
  labs(x = "Labour time (person-hours)", y = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "top",
    legend.justification = "left",
    legend.key.size = unit(0.4, "cm"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(colour = "#52514e"),
    axis.text.y = element_text(colour = "#0b0b0b", size = 10)
  )

group_hours <- tapply(task_summary$hours, task_summary$group, sum)
```

Grouped this way, the tasks that directly empty the pit (fluidizing,
extraction, sludge handling and disposal) account for
`r fmt(group_hours[["Emptying the pit"]])` person-hours
(`r fmt(100 * group_hours[["Emptying the pit"]] / total_labour_h)` %), about the
same as Rest (`r fmt(group_hours[["Rest"]])` person-hours). Support tasks take
`r fmt(group_hours[["Support tasks"]])` person-hours
(`r fmt(100 * group_hours[["Support tasks"]] / total_labour_h)` %) and site and
structure works `r fmt(group_hours[["Site and structure works"]])` person-hours
(`r fmt(100 * group_hours[["Site and structure works"]] / total_labour_h)` %).

@fig-jobs shows the labour time of each job. Labour time varies widely between jobs;
@tbl-jobs in the appendix lists the values for every job.

```{r}
#| label: fig-jobs
#| fig-cap: "Labour time (person-hours) per pit emptying job, ordered by date."
#| fig-width: 6.5
#| fig-height: 6
jobs |>
  mutate(job_id = factor(job_id, levels = rev(job_id))) |>
  ggplot(aes(x = labour_h, y = job_id)) +
  geom_col(fill = "#2a78d6", width = 0.7) +
  geom_vline(xintercept = mean(jobs$labour_h), linetype = "dashed", colour = "#52514e") +
  annotate("text", x = mean(jobs$labour_h), y = n_jobs + 0.6,
           label = paste0("mean ", fmt(mean(jobs$labour_h)), " h"),
           hjust = -0.05, size = 3, colour = "#52514e") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  coord_cartesian(clip = "off") +
  labs(x = "Labour time (person-hours)", y = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(colour = "#52514e"),
    plot.margin = margin(12, 8, 4, 4)
  )
```

# Conclusion

The total amount of time spent on the `r n_jobs` observed pit emptying jobs was
**`r fmt(total_labour_h)` person-hours** of recorded labour
(`r fmt(total_elapsed_h)` elapsed clock hours). For TDABC purposes the
person-hour figure is the relevant quantity, since it reflects the labour
capacity consumed by the service.

# Appendix A: Job-level results {.appendix}

```{r}
#| label: tbl-jobs
#| tbl-cap: "Workers, labour time and elapsed time for each job."
jobs |>
  transmute(
    Job = job_id,
    Gulpers = n_workers,
    Entries = n_entries,
    `Labour (person-h)` = labour_h,
    `Elapsed (h)` = elapsed_h
  ) |>
  bind_rows(tibble(
    Job = "Total", Gulpers = n_workers, Entries = n_entries,
    `Labour (person-h)` = total_labour_h, `Elapsed (h)` = total_elapsed_h
  )) |>
  kable(digits = 1, format.args = list(big.mark = ","))
```

# Appendix B: Relevant literature {.appendix}

The five references in the author's Zotero library most relevant to this
dataset, with one sentence on why each matters here:

1. @yesaya2021cost apply time-driven activity-based costing to compost
   production in Blantyre, Malawi, making it the closest methodological
   precedent for turning task-level timing data like these into activity costs.
2. @wilcox2026formal analyse the costs of 23 formal manual emptying providers in
   Malawi and Uganda and find labour is what makes manual emptying expensive,
   which is why the person-hours measured here are the key input for costing.
3. @chipeta2017designing designed and tested the pedal-powered Gulper pump in
   Mzuzu and estimated that a pit can be emptied in 1–2 hours when trash
   content is low, a benchmark for the job durations recorded here.
4. @chiposa2017characterization characterise pit latrines in peri-urban Mzuzu,
   reporting that 90 % of households also dispose of domestic waste in their
   pits, which helps explain the time spent on tasks such as pulling out trash.
5. @burt2019costs estimate the cost of delivering pit emptying services in
   Kigali, Rwanda, at different scales, giving a comparison point for costs
   derived from the labour time reported here.

# References

::: {#refs}
:::
